---
title: "Mapping rates of homelessness in the United States"
author: "[Laurel  Boyd](https://github.com/laurelboyd)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Homelessness in the United States: How does King County compare?

This analysis attempts to answer the question of how seattle/King County is doing relative to the rest of the country in addressing homelessness, using only publically available data.

# Point In Time Counts

Point in time (PIT) surveys are conducted on a rrgulat basis throughout the US to estimate numbers of homeless individuals. Although required by the United States Department of Housing and Urban Development (HUD) for receipt of monies to provide homeless services by Continuity of Care (CoC) Regions, comparison between regions is limited due to the various different methodologies for conducting PIT surveys.[^1]  Without access to other data sources (we could think about hospitalization, incarceration, and other data collected by social service agencies), it is a reasonably proxy for assessing relative differences in homelessness throughout the country and seeing how King County compares.

For this analysis, we'll use the most recent (2017) Census population estimates from the American Community Survey.

[^1] https://files.hudexchange.info/resources/documents/PIT-Count-Methodology-Guide.pdf

This is a multi-step process (note that bullet 4 seems exceedingly simple but isn't!):

* Find 2017 Point in Time counts for the entire nation by CoC region
* Find HUD CoC shapefiles 
* Find county population estimates from most recent Census (American Community Survey estimates from 2017)
* Overlay CoC regions on top of census tracts and calculate population for each region. (Unininsured would be interesting to look at if there's time.)
* Divide homeless counts by total population to calculate homeless population rates.
* Map findings.


```{r load packages}

library(dplyr) #data manipulation
library(rio) #speedy data import tool
library(leaflet) #mapping package
library(mapview) #mapping package (leaflet wrapper)
library(httr) #need this to use "GET" call for getting CoC geodatabase
library(rgdal) # mapping statistics/functions
library(sf) # newer "spatial features" package
library(sp) # older spatial package; has some functionality the sf package doesn't have (e.g., "over")
library(ggplot2) #another way to make maps 
library(janitor) #cleaning, tables with totals
library(stringr) #regex functionality in dplyr
library(tidycensus) #Census population estimates
library(rgeos) #polygon to polygon intersections
library(RColorBrewer) #color palatte for choropleth maps
library(htmltools) #labels for leaflet
library(scales) #needed for comma function in leaflet legend
library(lwgeom)#fix geography with st_make_valid
```
```{r }
###################################################################################
####################  Import PIT counts ###########################################

#(Craggy note: This portion would ideally come from the PIT code in the repository - but it didn't work for me so I just grabbed the portion I need for mapping)

fieldmap <- rio::import(system.file("extdata", "pit_counts_coc_fieldmap_2007t2018.xlsx", package = "craggy2019"))

file <- system.file("extdata", "2007-2018-PIT-Counts-by-CoC.xlsx", package = "craggy2019")
y2017 <- rio::import(file, which = 2)[,fieldmap$y2017] #207 since that's the most recent year of Census estimates...
names(y2017) <- fieldmap$shortname
y2017$year <- 2017

#400 regions in the file including 2 that are bogus: 1 NA/Totals column and 1 gobbly gook: "a MO-604 covers territory in both Missouri..."
#y2017 %>% distinct(coc_number)

#drop down to 398
y2017 <-y2017 %>%
  filter(!str_detect(coc_number, 'a MO') & (!is.na(coc_number)))

#need numeric columns for joining to CoC regions
y2017$overall_homeless <- as.numeric(as.character(y2017$overall_homeless))
y2017$unsheltered_homeless <- as.numeric(as.character(y2017$unsheltered_homeless))
#drop everything else...
y2017 <- y2017[ -c(4:7, 9:28) ]

#look at distribution of homelessness across regions


#highlight king county
#y2017<-y2017 %>% mutate( ToHighlight = ifelse( y2017$coc_number == "WA-500", "yes", "no" ) )

#bar chart
# ggplot( y2017, aes( x = reorder(coc_name, overall_homeless), y = overall_homeless, fill = ToHighlight ) ) +
#     geom_bar( stat = "identity" ) +
#     scale_fill_manual( values = c( "yes"="orange", "no"="gray" ), guide = FALSE )+
#    ylim(0, 80000)+
#     xlab("308 Continuity of Care (CoC) Regions") +
#   ylab("Overall homeless count") +
#   ggtitle("Homeless counts across the United States, 2017")+
#   theme(axis.text.x=element_blank())+
#   scale_y_continuous(position = "right")

#there are more elegant ways of doing the below...
y2017$king <- 11643
y2017$LA <-52442
y2017$NYC <-76501
  
#histogram
ggplot(y2017, aes(overall_homeless))+
  geom_histogram(fill ="orange", color = "white", bins = 100) +
   xlim(0, 80000)+
  xlab("Overall homeless persons") +
  ylab("Number of regions") +
  ggtitle("Seattle is third in the nation in overall homeless counts for 2017")+
theme(plot.title = element_text(hjust = 0.5)) +
geom_vline(aes(xintercept = king), lty="dashed", color = "gray") +
  geom_vline(aes(xintercept = LA), lty="dashed", color = "gray") +
  geom_vline(aes(xintercept = NYC), lty="dashed", color = "gray") +
geom_text(aes(label  = paste("Seattle = ", king), x= king, y = 55))+
geom_text(aes(label  = paste("Los Angeles = ", LA), x= LA, y = 55))+
geom_text(aes(label  = paste("NYC = ", NYC), x= NYC, y = 55))   + theme_minimal()

#drop extra vars
y2017 <- y2017[ -c(5:9) ]
```

The Seattle/King County CoC is third in the nation in terms of total overall homeless. In looking at the histogram of homeless counts by homeless (CoC) region, we can see that most regions have 10,000 or fewer homeless people (indeed, half of all regions have 549 or fewer homeless persons), but that several regions (Seattle, Los Angeles and New York City) have many more homeless. 

Next we'll look at the rate of homelessness by region.

```{r }
###################################################################################
####################  Import CoC regions and PIT counts#######################

# HUD input file geodatabase
url <- "https://www.hudexchange.info/resource/coc/gis/CoC_GIS_NatlTerrDC_Shapefile_2017.zip"
GET(url, write_disk("data.zip", overwrite = TRUE))
unzip("data.zip", overwrite = TRUE) #unzip file from downloads

# List all feature classes in a file geodatabase (only one feature class in this geodatabase)
#ogrListLayers("FY17_CoC_National_Bnd.gdb")

#Check out what format the data is in (should be SpatialPolygonsDataFrame)
#class(FY17COC)

# Read in the feature class
x <- st_read(dsn="FY17_CoC_National_Bnd.gdb")

# Rename column names in geodatabase layer to match PIT counts file
names(x)[names(x) == 'COCNUM'] <- 'coc_number'
names(x)[names(x) == 'COCNAME'] <- 'coc_name'

#join 2017 PIT counts to map before converting to a spatal layer
#y <- dplyr::inner_join(x, y2017)
FY17COC <- sf::st_as_sf(x) #convert to "sf" type
#class(FY17COC) #yep, now "sf"
#need to define datum for use in later operations
FY17COC <- st_transform(FY17COC, 4326) 
#sf::st_crs(FY17COC) #check coordinate reference system - is the datum defined? yep.

#map CoC regions
#mapview(FY17COC)

 ```

```{r }
###################################################################################
####################  Import census data and join layers ##########################

#load population counts fron tidycensus package (most recent American Community Survey data is from 2017)
#http://zevross.com/blog/2018/10/02/creating-beautiful-demographic-maps-in-r-with-the-tidycensus-and-tmap-packages/

#step 1: request Census API: https://api.census.gov/data/key_signup.html
#step 2: figure out which table (total population counts) to import (e.g. "B######")-> go to  https://censusreporter.org/
#step 3: clean geometries
#step 4: convert counties to centroids
#step 5: add buffer to centroids (see side note below for reason)
#step 6: join county centriods to CoC regions a
#step 7: note the large file size from step 6: silently dissolve counties into regions 
#step 8: merge PIT counts to spatial layer an calculate homelessness rates at CoC region level
#step 9: map findings

#STEP 1:
#census_api_key("YOUR KEY GOES HERE", install = TRUE)
#readRenviron("~/.Renviron")#run this the first time you use this key
#Sys.getenv("CENSUS_API_KEY")#Check yer work!

#STEP 2:

 COUNTIES <- get_acs("county", table= "B00001", year = 2017,
 output = "tidy", state = NULL, geometry = TRUE, shift_geo = FALSE) %>% #add cache_table = TRUE first time you run this
  rename (`county_population` = estimate) 
COUNTIES <- st_transform(COUNTIES, 4326) #need to define datum for merging - and use by leaflet

#STEP 3:
#check to see if the geometries are valid -> #https://www.r-spatial.org/r/2017/03/19/invalid.html
#turns out the CoC shapefile is not valid...
#st_is_valid(FY17COC) #nope; use some tools below when you join to fix this issue

#STEP 4:
#convert counties to centroids (waaay less processing time than joining polygons to polygons; fixes many issues where the polygon shapes don't match...)

#references: https://r-spatial.github.io/sf/reference/st_join.html

        # # #case study/demo: WA state
        # # 
        # #WA counties only
        # WACOUNTIES <-COUNTIES %>%
        #   filter(str_detect(GEOID, "^53"))
        # 
        # #WA CoC regions only
        # WACOC <-FY17COC %>%
        #   filter(str_detect(STATE_NAME, "^Washing"))
        # 
        # #select counties cooresponding to "WA balance of State CoC"
        # restofWA <-WACOUNTIES%>%
        #   filter(!NAME %in%  c("King County, Washington", "Spokane County, Washington", "Yakima County, Washington", "Pierce County, Washington", "Snohomish County, Washington", "Clark County, Washington"))
        # 
        # #how many people live in this CoC?
        # restofWA%>%
        #   mutate(totalpop = sum(population))
        # 
        # #how many people live in all of WA?
        # WACOUNTIES%>%
        #   mutate(totalpop = sum(population))
        # 
        # mapview(restofWA)+mapview(AAA1)
        # 
        # #so, what happens when we join the centroids to CoC regions?
        # AAA1 <-st_join(st_make_valid(WACOC), WACOUNTIES, left = TRUE) %>%
        #   group_by(coc_name)%>%
        #   mutate(region_population = sum(county_population)) %>%
        #   mutate(prop_homeless = overall_homeless/region_population) %>%
        #        ungroup()
        # 

        # mapview(AAA1)
        # AAA1$prop_homeless
        
        # #195064 total pop in "balance of state CoC"- 165927 population as calculated = 29137 difference!
        # #Turns out this is the sum of the populations of the island counties!
        # #centroids for these bad boys are somewhere in the ocean; they don't match up with the CoC regions...

#ahem - back to step 4: convert to centroids
COUNTIEScentroid <- st_centroid(COUNTIES) 
#step 5: add 1/10 mile buffer to centroids 
COUNTIEScentroid_buffer <-st_buffer(COUNTIEScentroid, 0.1) 
#step 6:join centroids to CoC regions 
FY17COC_COUNTIES <-st_join(st_make_valid(FY17COC), COUNTIEScentroid_buffer, left = FALSE)  #st_make_valid fixes issues in the CoC layer

#step 7: silently dissolve counties into regions and aggregate population counts
pop_by_regions = FY17COC_COUNTIES %>% group_by(coc_name) %>%
  summarize(pop = sum(county_population, na.rm = TRUE))

#step 8:add PIT attributes to regional sf object
percent_by_regions <- dplyr::inner_join(pop_by_regions, y2017)%>%
   mutate(prop_homeless = overall_homeless/pop) 

#look at the map
#mapview(percent_by_regions)


```
```{r }
###################################################################################
####################  Mapping rates ###############################################

# helpful resource: https://rstudio.github.io/leaflet/choropleths.html


#create bins for map
#distribution is left-skwewed so instead of using equal-sized bins (e.g. 10,000 people/bin) use quintiles instead

# FY17COC<- FY17COC %>%
#   filter(!is.na(overall_homeless))
# 
# qpal <- colorQuantile("Oranges", FY17COC$overall_homeless, n = 5)
# qpal_colors <- unique(qpal(sort(FY17COC$overall_homeless) )) # hex codes
# qpal_labs <- quantile(FY17COC$overall_homeless, seq(0, 1, .2)) # depends on n from pal
# qpal_labs <- paste(lag(qpal_labs), qpal_labs, sep = " - ")[-1] # first lag is NA


#map
# leaflet(FY17COC) %>%
#   addProviderTiles("Stamen.TonerLite") %>%
#    setView(-96, 37.8, 4)%>%
#    addPolygons(
#     # fill
#     fillColor   = ~qpal(overall_homeless),
#     fillOpacity = 0.7,
#     # line
#     dashArray   = "3",
#     weight      = 2,
#     color       = "white",
#     opacity     = 1,
#     # interaction
#     highlight = highlightOptions(
#       weight = 5,
#       color = "#666",
#       dashArray = "",
#       fillOpacity = 0.7,
#       bringToFront = TRUE),
#   label = labels,
#   labelOptions = labelOptions(
#     style = list("font-weight" = "normal", padding = "3px 8px"),
#     textsize = "15px",
#     direction = "auto")) %>%
#   addLegend(
#    colors = qpal_colors, labels = qpal_labs, opacity = 0.7, title = HTML("Total homelessness counts"),
#   position = "bottomright")

percent_by_regions$prop_homeless10k <- percent_by_regions$prop_homeless*10000
#some HMTL code to format labels
labels <- sprintf(
  "<strong>%s</strong><br/> Overall homeless per 10,000 people: %s",
  percent_by_regions$coc_name , comma(percent_by_regions$prop_homeless10k)) %>% 
  lapply(HTML)

binpal <- colorBin("Oranges", percent_by_regions$prop_homeless10k, 7, pretty = FALSE)


leaflet(percent_by_regions) %>%
  addProviderTiles("Stamen.TonerLite") %>%
   setView(-96, 37.8, 4)%>%
   addPolygons(
    # fill
    fillColor   = ~binpal(prop_homeless10k),
    fillOpacity = 0.7,
    # line
    dashArray   = "3",
    weight      = 2,
    color       = "white",
    opacity     = 1,
    # interaction
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
  label = labels,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>%
  addLegend(
   pal = binpal, values = ~prop_homeless10k, opacity = 0.7, title = HTML("Overall homelessness rates per 10,000 population"),
  position = "bottomright")

#histogram
percent_by_regions$king <- 778
ggplot(percent_by_regions, aes(prop_homeless10k))+
  geom_histogram(fill ="orange", color = "white", bins = 100) +
   #xlim(0, 80000)+
  xlab("Rate of homeless persons per 10,000 population") +
  ylab("Number of regions") +
  ggtitle("Seattle is tenth in the nation in rate of overall homeless for 2017")+
theme(plot.title = element_text(hjust = 0.5))+
geom_vline(aes(xintercept = king), lty="dashed", color = "gray") +
 # geom_vline(aes(xintercept = LA), lty="dashed", color = "gray") +
  #geom_vline(aes(xintercept = NYC), lty="dashed", color = "gray") +
geom_text(aes(label  = paste("Seattle = ", king), x= king, y = 30))
#+
#geom_text(aes(label  = paste("Los Angeles = ", LA), x= LA, y = 55))+
#geom_text(aes(label  = paste("NYC = ", NYC), x= NYC, y = 55))   + theme_minimal()

```

## How does Seattle CoC compare with other areas of the US?

From mapping overall rates of homeless by population, we can see that the West Coast (San Francisco, Oregon)
